Cosmology with Weak Lensing photometric surveys: constraints with redshift tomography of non--Gaussian statistics

### 2016-04-06 photoz_paper/paper.tex
finished implementing Zoltan suggestions

\begin{abstract}
Weak Gravitational Lensing is becoming a popular technique to constrain cosmological parameters, such as the Dark Energy equation of state $w$. When analyzing galaxy surveys, redshift information has proven to be a valuable addition to angular shear correlations. We forecast parameter constraints on the triplet $(\Omega_m,w,\sigma_8)$ for a LSST like galaxy photometric survey, using tomography of the shear-shear power spectrum, convergence peak counts and higher convergence moments. We find that the addition of peak and counts to the power spectrum can significantly improve the constraints on the $(\Omega_m,w)$ doublet. We evaluate the effect of uncorrected photometric redshift systematics on parameter constraints. We find that different statistics lead to different bias directions in parameter space, leaving a window open for self--calibration. 
\end{abstract}

@@

The measured galaxy ellipticity $\pmb{\epsilon}$ is an estimate of the cosmic shear $\pmb{\gamma}$ due to Large Scale Structure if the non--lensed galaxy shape is a circle. If this assumption is relaxed, the measured galaxy ellipticity $\pmb{\epsilon}_{\rm m}$ can be modeled as a cosmic shear term plus a noise term \citep{wlreview}
\begin{equation}
\pmb{\epsilon}_{\rm m} = \pmb{\gamma} + \pmb{\epsilon}_{\rm n}
\end{equation} 
%
where $\pmb{\epsilon}_{\rm n}$ is a random Gaussian variable with zero mean and redshift dependent variance $\sigma_n(z)=0.15+0.035z$. This is equivalent to say that the cosmic shear inferred from ellipticity observations $\pmb{\gamma}_{\rm m}$ can be written as the sum of the true cosmic shear plus a noise term $\pmb{\gamma}_{\rm n}$ with the same statistical properties as $\pmb{\epsilon}_{\rm n}$. 
We add independent random realizations of the shape noise $\pmb{\gamma}_{\rm n}$ to each of the $N_r$ shear catalogs. Each shape noise realization is generated with a different random seed. The same random seeds are used to generate shape noise catalogs across simulations with different cosmological parameters $\{\bb{p}_i\}$.

@@

\section{Results}
In this section we illustrate the main results of this work. Figure \ref{fig:pcacomponents} shows the robustness of the dimensionality reduction technique we adopted. on the three summary statistics considered, namely the convergence power spectrum $P^{\kappa\kappa}(\ell,\bar{z}_b,\bar{z}_{b'})$, peak counts $n_{\rm pk}(\nu,\bar{z}_b)$ and moments $\pmb{\mu}(\bar{z}_b)$. To measure the power spectrum we chose 15 uniformly spaced multipole bands between $(\ell_{\rm min},\ell_{\rm max})=(200,2000)$. Because, of the combinations $(\bar{z}_b,\bar{z}_{b'})$, only 15 of them (5 diagonal + 10 off-diagonal) are independent, this leads to a total of $N_b=15\times 15=225$ power spectrum measurement bands, including cross redshift information. 
We bin the convergence peak counts in 30 uniformly spaced $\nu$ bins between $(\nu_{\rm min},\nu_{\rm max})=(-2,7)$, for a total of $N_b=30\times5=150$ measurement bands, including redshift information. The total size of the moments summary statistic vector is $N_b=9\times 5=45$, including redshift information.

The forecast error bars on $w$ are calculated according to equation (\ref{meth:parcovestimator}), where the covariance matrix $\bbh{C}$ and its inverse $\bbh{\Psi}$ have been estimated with $N_r=16000$ realizations of each summary statistic in the fiducial cosmology. Figure \ref{fig:constraintsOm-w} shows the $1\sigma$ confidence contours on the $(\Omega_m,w)$ doublet calculated with equation (\ref{meth:parcovestimator}) after the PCA dimensionality reduction performed according to equation (\ref{meth:pcaprojection}), for a variety of choices $({\rm statistic},N_c)$. We also show the improvements on the confidence contours when combining different summary statistics after the corresponding dimensionality reductions have been performed.

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/photoz_bias_Om-w.eps}
\caption{Distribution of parameter estimates in the $(\Omega_m,w)$ parameter space using the power spectrum (red), peak counts (green) and moments (blue) to fit 20 LSST-like mock observations. The parameter deviations $(\delta\Omega_m,\delta w)$ are obtained comparing parameter estimates according to equation (\ref{meth:parestimate}) with and without photo-$z$ errors. Results for both optimistic (dots) and pessimistic (crosses) are shown. The colored squares and the ellipses correspond to respectively to the mean and $1\sigma$ level of the $(\delta\Omega_m,\delta w)$ distribution in the pessimistic systematics case, assuming this distribution is Gaussian.}
\label{fig:photozbias}
\end{figure}

@@

\section{Discussion}
In this section we discuss our findings. Figure \ref{fig:pcacomponents} shows that our dimensionality reduction technique is robust. For all the summary statistics we consider, we notice that a plateau in the $w$ error is reached for a sufficient high number of principal components $N_c$. We also see that for single redshift statistics, this plateau is reached for $\sim 5$ components for the power spectrum and the moments, and $\sim 10$ components for the peak counts. The number of components to consider increases when including redshift information, and can reach $\sim 30$ for the power spectrum and moments and $\sim 40$ for the peak counts. We can also see that most of the cosmological information is concentrated in the last redshift bin, with tomography improving the constraints by a factor of few for the power spectrum and peak counts.

In Figure \ref{fig:constraintsOm-w} we can see that peak counts and moments can significantly improve the constraints on $(\Omega_m,w)$ when combined with the power spectrum. Combining the peaks or the moments to the power spectrum improves the $w$ constraint by a factor of 1.7 while considering the three statistics together the error improves by an additional $30\%$. This shows that peaks and moments contain cosmological information that is not contained in the power spectrum, because a similar improvement cannot be obtained by increasing the number of PCA components in the power spectrum dimensionality reduction procedure.

@@

\section{Conclusions}
In this work we study cosmological parameter constraints forecasts for an LSST-like galaxy survey using the convergence power spectrum and a range of non--Gaussian statistics. We make use of redshift tomography to improve the constraints from their single redshift counterparts. We also investigate the effects of uncorrected photo-$z$ systematic effects on the inferred cosmology. We find that:
\begin{itemize}
	\item Principal Component analysis is a robust technique to keep the dimensionality of the parameter space under control and avoid the numerical pitfalls explained in \citep{Taylor12,DodelsonSchneider13,Taylor14} and more recently \citep{PetriVariance}. In particular we find that only a few components are necessary to characterize the cosmological information content in single redshift statistics, while more components are necessary when tomography is included. 
	\item Peak counts and moments can significantly improve $(\Omega_m,w)$ parameter constraints when used in combination with the power spectrum.
	\item Uncorrected photo-$z$ systematics can bias parameter constraints, but the direction of the bias is different for each summary statistic, leaving open windows for self-calibration
\end{itemize}

>>>>

\begin{abstract}
Weak Gravitational Lensing is becoming a popular technique to constrain cosmological parameters, such as the Dark Energy equation of state $w$. When analyzing galaxy surveys, redshift information has proven to be a valuable addition to angular shear correlations. We forecast parameter constraints on the triplet $(\Omega_m,w,\sigma_8)$ for a LSST like galaxy photometric survey, using tomography of the shear-shear power spectrum, convergence peak counts and higher convergence moments. We find that most the cosmological information is contained in the highest redshift galaxies, with redshift tomography further improving the $w$ constraints by only a few $10\%$. We also find that the addition of peak and counts to the power spectrum can significantly improve the constraints on the $(\Omega_m,w)$ doublet. We evaluate the effect of uncorrected photometric redshift systematics on parameter constraints. We find that different statistics lead to different bias directions in parameter space, leaving a window open for self--calibration. 
\end{abstract}

@@

The measured galaxy ellipticity $\pmb{\epsilon}$ is an estimate of the cosmic shear $\pmb{\gamma}$ due to Large Scale Structure if the non--lensed galaxy shape is a circle. If this assumption is relaxed, the measured galaxy ellipticity $\pmb{\epsilon}_{\rm m}$ can be modeled as a cosmic shear term plus a noise term \citep{wlreview}
\begin{equation}
\pmb{\epsilon}_{\rm m} = \pmb{\gamma} + \pmb{\epsilon}_{\rm n}
\end{equation} 
%
where $\pmb{\epsilon}_{\rm n}$ is a random Gaussian variable with zero mean and redshift dependent variance $\sigma_n(z_s)=0.15+0.035z_s$. This is equivalent to say that the cosmic shear inferred from ellipticity observations $\pmb{\gamma}_{\rm m}$ can be written as the sum of the true cosmic shear plus a noise term $\pmb{\gamma}_{\rm n}$ with the same statistical properties as $\pmb{\epsilon}_{\rm n}$. 
We add independent random realizations of the shape noise $\pmb{\gamma}_{\rm n}$ to each of the $N_r$ shear catalogs. Each shape noise realization is generated with a different random seed. The same random seeds are used to generate shape noise catalogs across simulations with different cosmological parameters $\{\bb{p}_i\}$.

@@

\section{Results}

In this section we illustrate the main results of this work. Figure \ref{fig:pcacomponents} shows the robustness of the dimensionality reduction technique we adopted. on the three summary statistics considered, namely the convergence power spectrum $P^{\kappa\kappa}(\ell,\bar{z}_b,\bar{z}_{b'})$, peak counts $n_{\rm pk}(\nu,\bar{z}_b)$ and moments $\pmb{\mu}(\bar{z}_b)$. To measure the power spectrum we chose 15 uniformly spaced multipole bands between $(\ell_{\rm min},\ell_{\rm max})=(200,2000)$. Because, of the combinations $(\bar{z}_b,\bar{z}_{b'})$, only 15 of them (5 diagonal + 10 off-diagonal) are independent, this leads to a total of $N_b=15\times 15=225$ power spectrum measurement bands, including cross redshift information. 
We bin the convergence peak counts in 30 uniformly spaced $\nu$ bins between $(\nu_{\rm min},\nu_{\rm max})=(-2,7)$, for a total of $N_b=30\times5=150$ measurement bands, including redshift information. The total size of the moments summary statistic vector is $N_b=9\times 5=45$, including redshift information.

The forecast error bars on $w$ are calculated according to equation (\ref{meth:parcovestimator}), where the covariance matrix $\bbh{C}$ and its inverse $\bbh{\Psi}$ have been estimated with $N_r=16000$ realizations of each summary statistic in the fiducial cosmology.

Figure \ref{fig:nopca} shows a comparison between the $w$ constraints obtained using single redshift bins, with and without PCA dimensionality reduction, and compares these single redshift constraints with the ones obtained using redshift tomography.

\begin{figure*}
\includegraphics[scale=0.3]{Figures/w_power_spectrum_no_pca.pdf}
\includegraphics[scale=0.3]{Figures/w_peaks_no_pca.pdf}
\includegraphics[scale=0.3]{Figures/w_moments_no_pca.pdf}
\caption{Comparison between single redshift $w$ constraints (marginalized over $(\Omega_m,\sigma_8)$) obtained without PCA (gray shaded bars) and with PCA (colored bars) as a function of the redshift bin $\bar{z}_b$. The rightmost black bar shows the same constraint, but obtained with redshift tomography and using the maximum number of PCA components seen in Figure \ref{fig:pcacomponents}. We show constrained obtained with the power spectrum (top right), peak counts (top left) and moments (bottom).}
\label{fig:nopca}
\end{figure*}

Figure \ref{fig:constraintsOm-w} shows the $1\sigma$ confidence contours on the $(\Omega_m,w)$ doublet calculated with equation (\ref{meth:parcovestimator}) after the PCA dimensionality reduction performed according to equation (\ref{meth:pcaprojection}), for a variety of choices $({\rm statistic},N_c)$. We also show the improvements on the confidence contours when combining different summary statistics after the corresponding dimensionality reductions have been performed.

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/photoz_bias_Om-w.eps}
\caption{Distribution of parameter estimates in the $(\Omega_m,w)$ parameter space using the power spectrum (red), peak counts (green) and moments (blue) to fit 20 LSST-like mock observations. The parameter deviations $(\delta\Omega_m,\delta w)$ are obtained comparing parameter estimates according to equation (\ref{meth:parestimate}) with and without photo-$z$ errors. The colored squares and the ellipses correspond to respectively to the mean and $1\sigma$ level of the $(\delta\Omega_m,\delta w)$ distribution in the pessimistic systematics case, assuming this distribution is Gaussian.}
\label{fig:photozbias}
\end{figure}

@@

\section{Discussion}
In this section we discuss our findings. Figure \ref{fig:pcacomponents} shows that our dimensionality reduction technique is robust. For all the summary statistics we consider, we notice that a plateau in the $w$ error is reached for a sufficient high number of principal components $N_c$. We also see that for single redshift statistics, this plateau is reached for $\sim 5$ components for the power spectrum and the moments, and $\sim 10$ components for the peak counts. Moreover, Figure \ref{fig:nopca} shows that, at least for the four highest redshift bins $\{\bar{z}_b: b\geq 2\}$, most of the cosmological information contained in the full (pre-PCA) summary statistic vector can be captured with a limited number of principal components $N_c<N_b$.

The number of components to consider increases when including redshift information, and can reach $\sim 30$ for the power spectrum and moments and $\sim 40$ for the peak counts. Figure \ref{fig:nopca} clearly that most of the cosmological information is concentrated in the last redshift bin. The use of redshift tomography only tightens the constraints by a few $10\%$.

In Figure \ref{fig:constraintsOm-w} we can see that peak counts and moments can significantly improve the constraints on $(\Omega_m,w)$ when combined with the power spectrum. Combining the peaks or the moments to the power spectrum improves the $w$ constraint by a factor of 1.7 while considering the three statistics together the error improves by an additional $30\%$. This shows that peaks and moments contain cosmological information that is not contained in the power spectrum, because a similar improvement cannot be obtained by simply increasing the number of PCA components in the power spectrum dimensionality reduction procedure.

@@

\section{Conclusions}
In this work we study cosmological parameter constraints forecasts for an LSST-like galaxy survey using the convergence power spectrum and a range of non--Gaussian statistics. We make use of redshift tomography to improve the constraints from their single redshift counterparts. We also investigate the effects of uncorrected photo-$z$ systematic effects on the inferred cosmology. We find that:
\begin{itemize}
	\item Principal Component analysis is a robust technique to keep the dimensionality of the parameter space under control and avoid the numerical pitfalls explained in \citep{Taylor12,DodelsonSchneider13,Taylor14} and more recently \citep{PetriVariance}. In particular we find that only a few components are necessary to characterize the cosmological information content in single redshift statistics, while more components are necessary when tomography is included. Nevertheless we find that the number of required components $N_c$ is significantly smaller than the full summary statistic space dimensionality before performing PCA.
	\item Most of the cosmological information is contained in high redshift galaxies, with tomography improving $w$ constraints by only a few $10\%$.   
	\item Peak counts and moments can significantly improve $(\Omega_m,w)$ parameter constraints when used in combination with the power spectrum.
	\item Uncorrected photo-$z$ systematics can bias parameter constraints obtained from the power spectrum and the moments, but in different parameter directions, leaving open windows for self-calibration
\end{itemize}
